Learning Excursion Sets of Vector-valued Gaussian Random Fields for Autonomous Ocean Sampling

### 2021-03-12 main_web.tex
Updates from Overleaf

Monitoring the world's oceans has gained increased importance in light
of the changing climate and increasing anthropogenic impact. Central
to understanding the changes taking place in the upper water column is
knowledge of bio-geophysical interactions driven by an
agglomeration of physical forcings (e.g. wind, topography, bathymetry,
tidal influences) and incipient micro-biology driven by
planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers.  These often
result in a range of ecosystem-related phenomena such as blooms and
plumes, with direct and indirect effects on society \citep{ryan2017}. One of the
bottlenecks in the study of such phenomena, however, lies in the lack of
observational data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations in which ocean processes transpire, prompting the need for
effective means of data collection.  By \emph{sampling}, we refer here
primarily to the design of observational strategies in the spatial
domain with the aim to pursue measurements with high scientific
relevance. Models and methods from spatial statistics and experimental design can clearly contribute to this sampling challenge.

Data collection at sea has typically been based on static buoys,
floats, or ship-based methods, with significant logistical limitations
that directly impact coverage and sampling resolution. Modern methods
using satellite remote-sensing provide large-scale coverage but have
limited resolution, are limited to sensing the surface, and are
impacted by cloud cover. Numerical ocean models similarly find it
challenging to provide detail at fine scale \citep{Lermusiaux:2006},
and also come with computational costs that can be limiting. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean (Fig.~\ref{fig:envir1}). In particular,
autonomous underwater vehicles (AUVs) have advanced the state of data
collection and consequently have made robotics an integral part of
ocean observation \citep{das11b,Das2015,fossuminformation,fossum18b}.

@@

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. Missions can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations, making the
data collection \emph{adaptive}.  In doing so, the space of sampling
opportunities is still limited by a waypoint graph, which forms a
discretization of the search domain where the AUV can navigate;
however, the AUV can now modify its path at each waypoint based on
in-situ measurements and calculations onboard using onboard
deliberation \citep{py10,Rajan12,Rajan12b}.  Full numerical ocean
models based on complex differential equations cannot be run onboard
the AUV with limited computational capacity. Instead, statistical models
relying on random field assumptions are relevant as a means to
effectively update the onboard model from in-situ data, and to guide
AUV data collection trajectories.

The work presented here is primarily inspired by a case study
pertaining to using an AUV for spatial characterization of a frontal
system generated by a river plume. Fig.~\ref{fig:nidelven} shows the
survey area in Trondheim, Norway, where cold freshwater enters the fjord from a river, creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coriolis force, the
cold freshwater tends to flow east. Depending on the variations in
river discharge, tidal effects, coastal current and wind, this
boundary often gets distorted, and knowledge about its location is
highly uncertain, making deterministic planning challenging. The goal
is therefore to use AUV measurements for improved description of the
interface between fresh and oceanic waters, which in turn enables more direct sampling of biological dynamics. It is often not possible
to sample these biological variables directly, but off-the-shelf instruments provide temperature and
salinity measurements which serve as proxies for the underlying
biological phenomena. With the help of a vector-valued random field
model for temperature and salinity, one can then aim to describe the
plume. The goal of plume characterization will in this setting relate to that of estimating an excursion set (ES), which is implicitly defined over the vector-valued random field. 
In the context of environmental sampling, the joint salinity and
temperature excursions of a river plume help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}. Motivating examples for ESs of
multivariate processes are also abundant in other contexts, for
instance in medicine, where physicians do not rely solely on a single
symptom but must see several combined effects before making a
diagnosis.

@@

Surprisingly less attention has been dedicated to sequential
strategies in the case of vector-valued observations. It has been long
acknowledged that co-Kriging could be updated efficiently in the
context of sequential data assimilation \citep{Vargas-Guzman1999}, but
sequential strategies for estimating features of vector-valued random
fields are still in their infancy. \cite{LeGratiet.etal2015} use
co-Kriging-based sequential designs to multi-fidelity computer codes
and \cite{Poloczek2017} use related ideas for multi-information
source optimization, but not for ESs like we do here. More relevant
to our setting, \citet[p.82]{stroh} mentions general possibilities of
step-wise uncertainty reduction strategies for ESs in the context of
designing fire simulations, yet outputs are mainly assumed
independent.

@@

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Smith2016,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\citep{bueso1998state,zidek2019monitoring}. However, in GRF models, response variance and entropy only depend on measurement locations and not on response values. This means that there is limited flexibility for
active adaptation of trajectories based on measurement values. The use
of data-driven adaptive criteria was introduced to include more
targeted sampling of regions of scientific interest in \cite{Low2009}
and \cite{fossuminformation}.

@@

We denote by $\gp$ a vector-valued random field indexed by some
arbitrary domain $\domain$, and assume values of the field at any
fixed location $\x \in \domain$, denoted $\gp[\x]$, to be a
$\no$-variate random vector ($\no\geq 2$). In the river plume
characterization case, $\domain$ is a prescribed domain in
Trondheimsfjord, Norway (for the purpose of our AUV application, a
discretization of a $2$-dimensional domain at fixed depth is
considered), and $\no=2$ with responses of temperature and salinity. A
multivariate GRF model is assumed for $\gp$. To illustrate these concepts,
Fig.~\ref{fig:real_temp} and \ref{fig:real_sal} show a realization of such a
vector-valued GRF on $\domain=[0,1]^2$. Fig.~\ref{fig:jointex_roi} represents a by-product of interest derived from these
realizations, namely regions: i) in red, where both temperature and salinity are high (i.e., exceeding respective thresholds), indicative of ocean water, ii) in white, where both temperature and salinity are low, indicative of riverine water, and iii) in light-red, where one variable is above and the other below their respective thresholds, indicative of mixed waters.

\begin{figure}[!b] 
\centering 
\subfigure[Temperature]{\includegraphics[
height=0.27\textwidth,keepaspectratio]{Figures/intro_temp.png}\label{fig:real_temp}}
\subfigure[Salinity]{\includegraphics[
height=0.27\textwidth,keepaspectratio]{Figures/intro_sal.png}\label{fig:real_sal}}
\subfigure[Regions of interest]{\includegraphics[
height=0.27\textwidth,keepaspectratio]{Figures/intro_excu.png}\label{fig:jointex_roi}}
\caption{Realization of a bivariate GRF (display (a) and (b)) and ES above some threshold (c). Joint excursion in red and excursion of a single variable in light-red.}
\label{example_excu}
\end{figure}

@@

Assuming that $n$ batches of observations are available with sizes
$q_1,\dots, q_n$, and that one wishes to predict $\gp[\bm{x}]$ for
a batch of $q\geq 1$ generalized locations
$\bm{x}$, the simple co-Kriging mean then amounts to Kriging with respect to a
scalar-valued GRF indexed by $\domain\times \{1\dots,p\}$:
%
\begin{equation}\label{eq:cokrig_mean}
\mu_{[n]}(\bm{x})=\mu(\bm{x})+\lambda_{[n]}(\bm{x})^T (\mathbf{z}_{[n]}-\mu(\bm{x})).
\end{equation}
Here, $\mathbf{z}_{[n]}$ stands for the ($\sum_{i=1}^n
q_i$)-dimensional vector of observed (noisy) responses of
$Z$ at all considered generalized locations, and
$\lambda_{[n]}(\bm{x})$ is a vector of weights equal to
$$\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)^{-1} k(\bm{x}_{[n]}, \bm{x})
$$,
with $\bm{x}_{[n]}=(\bm{x}_1,\dots,
\bm{x}_n)$ and where
$\Delta_{[n]}$ is the covariance matrix of Gaussian-distributed noise
assumed to have affected measurements up to batch
$n$. For our applications with salinity and temperature observations,
this matrix is diagonal because we assume conditionally independent
sensor readings, but it might not be diagonal with other types of
combined measurements.  The matrix in parentheses will be assumed to
be non-singular throughout the presentation. The associated
co-Kriging %(conditional)
residual (cross-)covariance function can also be expressed in the same
vein via
%
\begin{equation}\label{eq:cokrig_cov}
k_{[n]}(\bm{x},\bm{x}')=k(\bm{x},\bm{x}')-\lambda_{[n]}(\bm{x})^T 
\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)
\lambda_{{[n]}}(\bm{x}').
\end{equation}

@@

\subsection{UQ on ESs of multivariate GRFs}
\label{sec:set_uq}

@@

Say we want to study the EP of a bivariate Gaussian
vector variable below some thresholds. The thresholds are set equal to the mean:
$\mu_1=t_1=5^o C$ for temperature and $\mu_2=t_2=30$ g/kg for
salinity. We play with the temperature and salinity correlation
and variances to study the effect on the EP and EBV.

@@

We now study how the EBV associated with data
collection at a point changes if only one of the two components of the
field is observed. We first draw a realization of the GRF defined
above and use it as ground-truth to mimic the real data-collection
process. A first set of observations are done at the locations
depicted in grey (see Fig.~\ref{fig:ebv_comp}), and the data is used
to update the GRF model. We then consider the green triangle as a
potential next observation location and plot the EBV reduction (at
each grid node in the waypoint graph) that would result from observing
only one component of the field (temperature or salinity), or both at
that point.

\begin{figure}[!b] 
\centering 
\subfigure[Regions of interest.]{\includegraphics[
height=0.21\textwidth,keepaspectratio]{Figures/ebv_comp_excu.png}\label{fig:ebv_comp_excu}}
\subfigure[Temperature.]{\includegraphics[
height=0.22\textwidth,keepaspectratio]{Figures/ebv_comp_temp.png}\label{fig:ebv_comp_temp}}
\subfigure[Salinity.]{\includegraphics[
height=0.22\textwidth,keepaspectratio]{Figures/ebv_comp_sal.png}\label{fig:ebv_comp_sal}}
\subfigure[Both.]{\includegraphics[
height=0.22\textwidth,keepaspectratio]{Figures/ebv_comp_both.png}\label{fig:ebv_comp_both}}
\caption{Pointwise Bernoulli variance reduction for observation of a
  single or both components of the random field at one location. Data
  collection locations in green. True excursion set in red. Places
  where only one response is above threshold are depicted in light-red. EBV
  reduction associated to observing one or both responses at the green
  location are shown in \ref{fig:ebv_comp_temp},
  \ref{fig:ebv_comp_sal} and \ref{fig:ebv_comp_both}.}
\label{fig:ebv_comp}
\end{figure}

Note that plotting the EBV reduction at each point might also be used
to compare different data collection plans. For example,
Fig. \ref{fig:ebv_north_vs_east} shows the EBV reduction associated
with a data collection plan along a vertical line (static north) and
one associated with a horizontal (static east). Both expectations are
computed according to the a-priori distribution of the GRF (i.e. no
observations have been included yet).

\begin{figure}[ht] 
\centering 
\subfigure[Excursion probability.]{\includegraphics[
height=0.25\textwidth,keepaspectratio]{Figures/ebv_static_excu.png}\label{fig:ebv_static_excu}}
\subfigure[Static north design.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_north.png}\label{fig:ebv_static_north}}
\subfigure[Static east design.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_east.png}\label{fig:ebv_static_east}}
\caption{Pointwise Bernoulli variance reduction for two different static designs (later noted as \textit{static\_north} and \textit{static\_east}). The prior EP is shown in
  \ref{fig:ebv_static_excu}. EBV reduction for each design shown in
  \ref{fig:ebv_static_north} and \ref{fig:ebv_static_east}.}
\label{fig:ebv_north_vs_east}
\end{figure}

@@

From a sequential point of view, $n$ data collection steps have
already been performed and one wants to choose what data to collect
next. The design evaluations
are based on the conditional expectation $\currentExp{.}$ from the law
of the field, conditional on all data available at
stage $n$.
Once the best design at stage $n$
has been selected, the data are collected and the GRF
model is updated using co-Kriging
Eq. \ref{eq:meanCoK} and \ref{eq:varCoK}, yielding a conditional law $\mathbb{P}_{[\stage+1]}$
after which the process is repeated.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more stage of measurements; it does not
anticipate what the subsequent designs might offer beyond the first
stage.  Based on the currently available data the myopic strategy
selects the location that leads to the biggest reduction in EIBV:
\begin{criterion}[Myopic]
The next observation location $\spatloc_{\stage + 1}$ is chosen among
the minimizers in $\candidates$ of the criterion: 
\begin{equation}\label{critSEQ}
     C_{\text{myopic}}(u) = \EIBV_{[\stage]}\left(\spatloc\right)
\end{equation}
\end{criterion}

@@

\begin{figure}[ht] 
\centering 
\subfigure[Excursion realization.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_myopic_excu.png}\label{fig:real_myopic}}
\hspace{2em}
\subfigure[BV reduction.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_myopic_ebv.png}\label{fig:route_myopic}}
\hspace{2em}
\subfigure[Expected reduction of BV.]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_myopic_radar.png}\label{fig:compass_myopic}}
\caption{Example run of the myopic strategy on a realisation of the GRF
model from \ref{sec:including_spatiality}. Reduction in Bernoulli variance
compared to the prior is shown in \ref{fig:route_myopic}, with past observation
locations in green and current AUV position in cyan. The expected IBV reduction
associated to data collection at neighbouring nodes of the current location is
shown in \ref{fig:compass_myopic}. The thick and light color indicates the node at $30 ^{\circ}$ to be the best possible choice.}
\label{fig:ebv_myopic}
\end{figure}

@@

We now extend the myopic strategy by considering two stages of
measurements, which is optimal in that it accounts consistently for
the expectations and minimizations in these two stages, but no anticipation beyond that.

The principle of
two-step look-ahead is to select the next observation location $\spatloc_{\stage + 1}$ that yields the biggest reduction in EIBV if
we were to (optimally) add one more observation after that again. In order to formalize this concept, we must
extend the notation for EIBV in the future (after observation
$\stage + 1$ has been made). We let $\currentEIBV(\cdot; u, y)$
denote the EIBV where expectations are taken conditional on the data
available at stage $n$ and on an additional observation $y$ at $u$ at stage $n+1$.

\begin{criterion}[2-step look-ahead]
      The next observation location $\spatloc_{\stage + 1}$ is chosen among the minimizers in $\candidates$ of the criterion
      \begin{align}\label{critLA}
          C_{\text{2-steps}}(u) &= \mathbb{E}_{[\stage]}\left[\min_{\spatloc' \in
                  \candidates(\spatloc)} \EIBV_{[\stage]}\left(\spatloc' ; \spatloc,
      Y\right)\right]
      \end{align}
      where $Y$ is the random data realization of $\gp_{\spatloc}$ according to its
      conditional law at step $\stage$ with the
      dependence of the set of candidates on the current location
      having been made explicit for the second stage of measurements.
    \end{criterion}

@@

We compare three different static designs denoted
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} (a version of \textit{static\_north} where with some east-west transitions in a zigzag pattern) with the three described sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. The static
AUV sampling paths are pre-scripted and cannot be altered.
For a fixed survey length, a closed-form expression for the EIBV is
available as in Proposition \ref{propo_eibv}. However, for the sequential
approaches this is not the case. For comparison, the properties are
therefore evaluated using Monte Carlo integration over several replicates
of realizations from the model while conducting simulated sequential
surveys for each one. An example of such a realization with a \textit{myopic}
strategy is shown in Fig.~\ref{fig:ebv_myopic}.

>>>>

Monitoring the world's oceans has gained increased importance in light
of the changing climate and escalating anthropogenic impact. Central
to understanding the changes taking place in the upper water column is
knowledge of bio-geophysical interactions driven by an
agglomeration of physical forcings (e.g. wind, topography, bathymetry,
tidal influences) and incipient micro-biology driven by
planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers.  These often
result in a range of ecosystem-related phenomena such as blooms and
plumes, with direct and indirect effects on society \citep{ryan2017}. One of the
bottlenecks in the study of such phenomena, however, lies in the lack of
observational data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations in which ocean processes transpire, prompting the need for
effective means of data collection.  By \emph{sampling}, we refer here
primarily to the design of observational strategies in the spatial
domain with the aim to pursue measurements with high scientific
relevance. Models and methods from spatial statistics and experimental design can clearly contribute to this sampling challenge.

Data collection at sea has typically been based on static buoys,
floats, or ship-based methods, with significant logistical limitations
that directly impact coverage and sampling resolution. Modern methods
using satellite remote-sensing provide large-scale coverage but have
limited resolution, are constrained to sensing the surface, and are
impacted by cloud cover. Numerical ocean models similarly find it
challenging to provide detail at fine scale \citep{Lermusiaux:2006},
and also come with computational costs that can be limiting. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean (Fig.~\ref{fig:envir1}). In particular,
autonomous underwater vehicles (AUVs) have advanced the state of data
collection and consequently have made robotics an integral part of
ocean observation \citep{das11b,Das2015,fossuminformation,fossum18b}.

@@

Surveys with AUVs are usually limited to observations along fixed
transects that are pre scripted in mission plans created manually by a
human operator. Missions can operate on a scale of a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations, making the
data collection \emph{adaptive}.  In doing so, the space of sampling
opportunities is still limited by a waypoint graph, which forms a
discretization of the search domain where the AUV can navigate;
however, the AUV can now modify its path at each waypoint based on
in-situ measurements and calculations onboard using onboard
deliberation \citep{py10,Rajan12,Rajan12b}.  Full numerical ocean
models based on complex differential equations cannot be run onboard
the AUV with limited computational capacity. Instead, statistical models
relying on random field assumptions are relevant as a means to
effectively update the onboard model from in-situ data, and to guide
AUV data collection trajectories.

The work presented here is primarily inspired by a case study
pertaining to using an AUV for spatial characterization of a frontal
system generated by a river plume. Fig.~\ref{fig:nidelven} shows the
survey area in Trondheim, Norway, where cold freshwater enters the fjord from a river, creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coriolis force, the
cold freshwater tends to flow east. Depending on the variations in river discharge, tidal effects, coastal current, and wind, this boundary often gets distorted. As a result, knowledge about its location is highly uncertain, making deterministic planning challenging. The goal is therefore to use AUV measurements for improved description of the
interface between fresh and oceanic waters, which in turn enables more direct sampling of biological dynamics. It is often not possible
to sample these biological variables directly, but off-the-shelf instruments provide temperature and
salinity measurements which serve as proxies for the underlying
biological phenomena. With the help of a vector-valued random field
model for temperature and salinity, one can then aim to describe the
plume. The goal of plume characterization will in this setting relate to that of estimating an excursion set (ES), which is implicitly defined over the vector-valued random field. 
In the context of environmental sampling, the joint salinity and
temperature excursions of a river plume help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}. Motivating examples for ESs of
multivariate processes are also abundant in other contexts, for
instance in medicine, where physicians do not rely solely on a single
symptom but must see several combined effects before making a
diagnosis.

@@

Surprisingly less attention has been dedicated to sequential
strategies in the case of vector-valued observations. It has been long
acknowledged that co-Kriging could be updated efficiently in the
context of sequential data assimilation \citep{Vargas-Guzman1999}, but
sequential strategies for estimating features of vector-valued random
fields are still in their infancy. \cite{LeGratiet.etal2015} use
co-Kriging-based sequential designs to develop multi-fidelity computer codes
and \cite{Poloczek2017} use related ideas for multi-information
source optimization, but not for ESs like we do here. More relevant
to our setting, \citet[p.82]{stroh} mentions general possibilities of
step-wise uncertainty reduction strategies for ESs in the context of
designing fire simulations, yet outputs are mainly assumed
independent.

@@

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Smith2016,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
 that focus largely on reducing
predictive variance or entropy \citep{singh2009efficient,Binney2013}. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\citep{bueso1998state,zidek2019monitoring}. However, in GRF models, response variance and entropy only depend on measurement locations and not on response values. This means that there is limited flexibility for
active adaptation of trajectories based on measurement values. The use
of data-driven adaptive criteria was introduced to include more
targeted sampling of regions of scientific interest in \cite{Low2009}
and \cite{fossuminformation}.

@@

We denote by $\gp$ a vector-valued random field indexed by some
arbitrary domain $\domain$, and assume values of the field at any
fixed location $\x \in \domain$, denoted $\gp[\x]$, to be a
$\no$-variate random vector ($\no\geq 2$). In the river plume
characterization case, $\domain$ is a prescribed domain in
Trondheimsfjord, Norway (for the purpose of our AUV application, a
discretization of a $2$-dimensional domain at fixed depth is
considered), and $\no=2$ with responses of temperature and salinity. A
multivariate GRF model is assumed for $\gp$. To illustrate these concepts,
Fig.~\ref{fig:real_temp} and \ref{fig:real_sal} show a realization of such a
vector-valued GRF on $\domain=[0,1]^2$. Fig.~\ref{fig:jointex_roi} represents a by-product of interest derived from these
realizations, namely regions: i) in red, where both temperature and salinity are high (i.e., exceeding respective thresholds), indicative of ocean water, ii) in white, where both temperature and salinity are low, indicative of riverine water, and iii) in pink, where one variable is above and the other below their respective thresholds, indicative of mixed waters.

\begin{figure}[!b] 
\centering 
\subfigure[Temperature]{\includegraphics[
height=0.27\textwidth,keepaspectratio]{Figures/intro_temp.png}\label{fig:real_temp}}
\subfigure[Salinity]{\includegraphics[
height=0.27\textwidth,keepaspectratio]{Figures/intro_sal.png}\label{fig:real_sal}}
\subfigure[Regions of interest]{\includegraphics[
height=0.27\textwidth,keepaspectratio]{Figures/intro_excu.png}\label{fig:jointex_roi}}
\caption{Realization of a bivariate GRF (display (a) and (b)) and ES above some threshold (c). Joint excursion in red and excursion of a single variable in pink.}
\label{example_excu}
\end{figure}

@@

Assuming that $n$ batches of observations are available with sizes
$q_1,\dots, q_n$, and that one wishes to predict $\gp[\bm{x}]$ for
a batch of $q\geq 1$ generalized locations
$\bm{x}$, the simple co-Kriging mean then amounts to Kriging with respect to a
scalar-valued GRF indexed by $\domain\times \{1\dots,p\}$:
%
\begin{equation}\label{eq:cokrig_mean}
\mu_{[n]}(\bm{x})=\mu(\bm{x})+\lambda_{[n]}(\bm{x})^T (\mathbf{z}_{[n]}-\mu(\bm{x})).
\end{equation}
Here, $\mathbf{z}_{[n]}$ stands for the ($\sum_{i=1}^n
q_i$)-dimensional vector of observed (noisy) responses of
$Z$ at all considered generalized locations, and
$\lambda_{[n]}(\bm{x})$ is a vector of weights equal to
$$\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)^{-1} k(\bm{x}_{[n]}, \bm{x}),
$$
with $\bm{x}_{[n]}=(\bm{x}_1,\dots,
\bm{x}_n)$ and where
$\Delta_{[n]}$ is the covariance matrix of Gaussian-distributed noise
assumed to have affected measurements up to batch
$n$. For our applications with salinity and temperature observations,
this matrix is diagonal because we assume conditionally independent
sensor readings, but it might not be diagonal with other types of
combined measurements.  The matrix in parentheses will be assumed to
be non-singular throughout the presentation. The associated
co-Kriging %(conditional)
residual (cross-)covariance function can also be expressed in the same
vein via
%
\begin{equation}\label{eq:cokrig_cov}
k_{[n]}(\bm{x},\bm{x}')=k(\bm{x},\bm{x}')-\lambda_{[n]}(\bm{x})^T 
\left(k(\bm{x}_{[n]}, \bm{x}_{[n]})+\Delta_{[n]} \right)
\lambda_{{[n]}}(\bm{x}').
\end{equation}

@@

\subsection{Uncertainty quantification on ESs of multivariate GRFs}
\label{sec:set_uq}

@@

Say we want to study the EP of a bivariate Gaussian
vector variable below some thresholds. The thresholds are set equal to the mean:
$\mu_1=t_1=5^o C$ for temperature and $\mu_2=t_2=30$ g/kg for
salinity. We alter the temperature and salinity correlation
and variances to study the effect on the EP and EBV.

@@

We now study how the EBV associated with data
collection at a point changes if only one of the two components of the
field is observed. We first draw a realization of the GRF defined
above and use it as ground-truth to mimic the real data-collection
process. A first set of observations are done at the locations
depicted in grey (see Fig.~\ref{fig:ebv_comp}), and the data is used
to update the GRF model. We then consider the green triangle as a
potential next observation location and plot the EBV reduction (at
each grid node in the waypoint graph) that would result from observing either one or both components of the field (temperature or salinity) at that point.

\begin{figure}[!b] 
\centering 
\subfigure[Regions of interest]{\includegraphics[
height=0.21\textwidth,keepaspectratio]{Figures/ebv_comp_excu.png}\label{fig:ebv_comp_excu}}
\subfigure[Temperature]{\includegraphics[
height=0.22\textwidth,keepaspectratio]{Figures/ebv_comp_temp.png}\label{fig:ebv_comp_temp}}
\subfigure[Salinity]{\includegraphics[
height=0.22\textwidth,keepaspectratio]{Figures/ebv_comp_sal.png}\label{fig:ebv_comp_sal}}
\subfigure[Both]{\includegraphics[
height=0.22\textwidth,keepaspectratio]{Figures/ebv_comp_both.png}\label{fig:ebv_comp_both}}
\caption{Pointwise Bernoulli variance reduction for observation of a
  single or both components of the random field at one location. Data
  collection locations in green. True excursion set in red. Places
  where only one response is above threshold are depicted in pink. EBV
  reduction associated to observing one or both responses at the green
  location are shown in \ref{fig:ebv_comp_temp},
  \ref{fig:ebv_comp_sal} and \ref{fig:ebv_comp_both}.}
\label{fig:ebv_comp}
\end{figure}

Note that plotting the EBV reduction at each point might also be used
to compare different data collection plans. For example,
Fig. \ref{fig:ebv_north_vs_east} shows the EBV reduction associated
with a data collection plan along a vertical line (static north) and
one associated with a horizontal (static east). Both expectations are
computed according to the a priori distribution of the GRF (i.e. no
observations have been included yet).

\begin{figure}[ht] 
\centering 
\subfigure[Excursion probability]{\includegraphics[
height=0.25\textwidth,keepaspectratio]{Figures/ebv_static_excu.png}\label{fig:ebv_static_excu}}
\subfigure[Static north design]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_north.png}\label{fig:ebv_static_north}}
\subfigure[Static east design]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_static_east.png}\label{fig:ebv_static_east}}
\caption{Pointwise Bernoulli variance reduction for two different static designs (later noted as \textit{static\_north} and \textit{static\_east}). The prior EP is shown in
  \ref{fig:ebv_static_excu}. EBV reduction for each design shown in
  \ref{fig:ebv_static_north} and \ref{fig:ebv_static_east}.}
\label{fig:ebv_north_vs_east}
\end{figure}

@@

From a sequential point of view, $n$ data collection steps have
already been performed and one wants to choose what data to collect
next. The design evaluations
are based on the conditional expectation $\currentExp{.}$ from the law
of the field, conditional on all data available at
stage $n$.
Once the best design at stage $n$
has been selected, the data are collected and the GRF
model is updated using co-Kriging
Eq. \ref{eq:meanCoK} and \ref{eq:varCoK}, yielding a conditional law $\mathbb{P}_{[\stage+1]}$,
after which the process is repeated.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more stage of measurements; it does not
anticipate what the subsequent designs might offer beyond the first
stage.  Based on the currently available data, the myopic strategy
selects the location that leads to the biggest reduction in EIBV:
\begin{criterion}[Myopic]
The next observation location $\spatloc_{\stage + 1}$ is chosen among
the minimizers in $\candidates$ of the criterion: 
\begin{equation}\label{critSEQ}
     C_{\text{myopic}}(u) = \EIBV_{[\stage]}\left(\spatloc\right)
\end{equation}
\end{criterion}

@@

\begin{figure}[ht] 
\centering 
\subfigure[Excursion realization]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_myopic_excu.png}\label{fig:real_myopic}}
\hspace{2em}
\subfigure[BV reduction]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_myopic_ebv.png}\label{fig:route_myopic}}
\hspace{2em}
\subfigure[Expected reduction of BV]{\includegraphics[
height=0.26\textwidth,keepaspectratio]{Figures/ebv_myopic_radar.png}\label{fig:compass_myopic}}
\caption{Example run of the myopic strategy on a realisation of the GRF
model from \ref{sec:including_spatiality}. Reduction in Bernoulli variance
compared to the prior is shown in \ref{fig:route_myopic}, with past observation
locations in green and current AUV position in cyan. The expected IBV reduction
associated to data collection at neighbouring nodes of the current location is
shown in \ref{fig:compass_myopic}. The thick and light colored line indicates the node at $30 ^{\circ}$ to be the best possible choice.}
\label{fig:ebv_myopic}
\end{figure}

@@

We now extend the myopic strategy by considering two stages of
measurements; this is optimal in that it accounts consistently for
the expectations and minimizations in these two stages, but does not anticipate beyond that.

The principle of
two-step look-ahead is to select the next observation location $\spatloc_{\stage + 1}$ that yields the biggest reduction in EIBV if
we were to (optimally) add one more observation. In order to formalize this concept, we must
extend the notation for EIBV in the future (after observation
$\stage + 1$ has been made). We let $\currentEIBV(\cdot; u, y)$
denote the EIBV where expectations are taken conditional on the data
available at stage $n$ and on an additional observation $y$ at $u$ at stage $n+1$.

\begin{criterion}[2-step look-ahead]
      The next observation location $\spatloc_{\stage + 1}$ is chosen among the minimizers in $\candidates$ of the criterion
      \begin{align}\label{critLA}
          C_{\text{2-steps}}(u) &= \mathbb{E}_{[\stage]}\left[\min_{\spatloc' \in
                  \candidates(\spatloc)} \EIBV_{[\stage]}\left(\spatloc' ; \spatloc,
      Y\right)\right],
      \end{align}
      where $Y$ is the random data realization of $\gp_{\spatloc}$ according to its
      conditional law at step $\stage$, with the
      dependence of the set of candidates on the current location
      having been made explicit for the second stage of measurements.
    \end{criterion}

@@

We compare three different static designs denoted
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} (a version of \textit{static\_north} with some east-west transitions in a zigzag pattern) with the three described sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. The static
AUV sampling paths are pre-scripted and cannot be altered.
For a fixed survey length, a closed-form expression for the EIBV is
available as in Proposition \ref{propo_eibv}. However, for the sequential
approaches this is not the case. For comparison, the properties are
evaluated using Monte Carlo integration over several replicates
of realizations from the model while conducting simulated sequential
surveys for each one. An example of such a realization with a \textit{myopic}
strategy is shown in Fig.~\ref{fig:ebv_myopic}.
